An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-02-03 main.tex
fixing problems
Updates from ShareLaTeX

Runtime Verification (RV) consists of analyzing execution 
traces using formal techniques. RV is scalable and 
therefore has a big potential for infusion in software 
engineering practices. It includes e.g. monitoring the 
execution of a system against properties formulated in 
Linear Temporal Logic (LTL). 
LTL offers a succinct notation for writing useful 
specification properties. However, it is limited in 
expressiveness in the propositional case, and several 
theoretic (non-operational) extensions have therefore been proposed. Furthermore, for many practical cases, there is a need to monitor properties 
that carry data, where one can use formalisms like 
first-order LTL. We show that 
first-order LTL has similar expressive
limitations as the propositional version. We suggest here 
two related extensions for increasing the expressive power: 
one for propositional LTL and one for 
first-order LTL. These extensions have a simple incremental operational 
semantics that is more suitable for RV than previously 
suggested extensions. We show that the 
propositional extension has the same expressiveness as the 
classical extensions, and demonstrate its
adoption for runtime verification of first-order safety 
properties. Finally, we expand the BDD-based runtime verification 
tool \dejavu{}  to support our extension and we perform some 
experiments.

@@

>>>>

Runtime Verification (RV) consists of analyzing execution 
traces using formal techniques. 
%RV is scalable and 
%therefore has a big potential for infusion in software 
%engineering practices. 
It includes monitoring the 
execution of a system against properties formulated in 
Linear Temporal Logic (LTL). 
LTL offers a succinct notation for writing useful 
specification properties. However, it is limited in 
expressiveness in the propositional case, and several 
theoretic (non-operational) extensions have therefore been proposed. Furthermore, for many practical cases, there is a need to monitor properties 
that carry data, where one can use formalisms like 
first-order LTL. We show that 
first-order LTL has similar expressive
limitations as the propositional version. We suggest here 
two related extensions for increasing the expressive power: 
one for propositional LTL and one for 
first-order LTL. These extensions have a simple incremental operational 
semantics that is more suitable for RV than previously 
suggested extensions. We show that the 
propositional extension has the same expressiveness as the 
classical extensions, and demonstrate its
adoption for runtime verification of first-order safety 
properties. Finally, we expand the BDD-based runtime verification 
tool \dejavu{}  to support our extension and perform some 
experiments.

@@

\klaus{Above: why set of subsets?}

### 2019-02-03 rule-ltl.tex
Updates from ShareLaTeX

We present first an extension to the propositional case that is based on using additional (auxiliary) propositions, not appearing in the execution, which summarize the past 
of the current execution. These propositions obtain their value in a state as a function of the prefix of the execution up and including that state through a past temporal formula. This can be seen as adding
summary states to the temporal specification, resulting in a specification that is between logic and finite automata representation. This extension, which we believe is conceptually simpler
and more intuitive than other suggested extensions, is shown to be
equivalent to the other common specification formalism:
B\"{u}chi automata, regular expressions and quantified LTL.

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL.
We present such an extension where additional (auxiliary) relations are
added. We show that this extension is more expressive than
first-order LTL, e.g., by repeating Wolper's classical algorithm
with data. On the other hand, we show that for the first-order version, this has less 
expressive power than adding quantification over relations, in distinction from the propositional case.

These extensions to the logics are in particular natural for
runtime verification of past (safety) temporal properties, since
the values of propositional and first-order LTL are based
on calculating and updating a summary of the prefix of the execution.
We provide runtime verification algorithms for the extended
logics and present an extension of the 
\dejavu{} tool (https://github.com/havelund/dejavu) that
realizes the extended first-order past temporal logic verification.
Finally, we provide some experimental results.

\klaus{I think that the structure of the paper should be
crystal clear so that the reader does not get lost. A lost reader leads to rejection. If a reader understand the structure, but perhaps not the details, the reader may still be happy.}

\subsubsection{Related Work}

For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds
state-dependent quantification over propositions to LTL.

@@

The extensions we suggest are in particular easy to implement, as an extension to existing propositional/first-order runtime verification. We extend our tool \dejavu, experiment with new examples that take advantage of the extension.
\fi

\section{Preliminaries}
\label{sec:prelim}

In this paper, we treat several versions of the logic LTL. To simplify
the presentation, we prefix LTL with letters as follows:

@@

\section{Extending Propositional LTL}
\label{sec:extending-prop-ltl}

@@

\section{Extending First-Order LTL}

@@

\begin{equation}
\label{EFLTL}
\psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n 
\end{equation}

>>>>

We present first an extension of the propositional LTL that is based on using additional (auxiliary) propositions, not appearing in the execution, which summarize the past 
of the current execution. These propositions obtain their value in a state as a function of the prefix of the execution up and including that state through a past temporal formula. This can be seen as adding
summary states to the temporal specification, resulting in a specification that is in between logic and finite automata representation. This extension, which we believe is conceptually simpler
and more intuitive than other suggested extensions, is shown to be
equivalent to the other common specification formalism:
B\"{u}chi automata, regular expressions and quantified LTL.

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL. Another example is based on the inexpressiveness of transitive closure of a relation in first order logic.
We present an extension where auxiliary relations, not appearing in the model, are
added. While this extension
is capable of expressing the above
properties, we show that for the first-order version, this has less 
expressive power than adding quantification over relations, in distinction from the corresponding propositional case.

These extensions to the logics are in particular natural for
runtime verification of past (safety) temporal properties, since
the values of propositional and first-order LTL are based
on calculating and updating a summary of the prefix of the execution. Runtime verification often concentrates on a past  version of temporal logic~\cite{HR} mainly because for these properties we can decide when the monitored property is violated immediately when this happen, while for future properties we may never know for sure. Consistent with that, 
we provide runtime verification algorithms for the past-time versions of the extended propositional and first-order
logics. We further present an extension of the 
\dejavu{} tool (https://github.com/havelund/dejavu) that
realizes the extended first-order past temporal logic verification. The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first order temporal logic over infinite domains (e.g., the integers, strings). It achieves efficiency by BDD representation of enumerations of data (rather than over the data values themselves) and using
a garbage collection algorithm. 
Finally, we provide some experimental results.

The contributions of this paper are both theoretical and practical. On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions. 
On the practical side, we present algorithms for the extended logic, and 
show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.

The structure of this paper is as follows. In Section~2 we first define the syntax and semantics of propositional temporal logic. We then review some of its classical extensions and present our own extension. In Section~3 we present a runtime verification algorithm for the extended past LTL. In Section~4 we review first order linear temporal logic, present our extension and study the relationships between the presented versions of first-order LTL. IIn Section~5 we present a runtime verification algorithm for the past version of the extended first-order LTL. Section~6 describes the implementation and provides experimental results. Finally, Section~7 concludes the paper.

%\klaus{I think that the structure of the paper should be
%crystal clear so that the reader does not get lost. A lost reader leads to rejection. If a reader understand the structure, but perhaps not the details, the reader may still be happy.}

\vspace{1ex}
\noindent {\bf Related Work}
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds
state-dependent quantification over propositions to LTL.

@@

The extensions we suggest are in particular easy to implement, as an extension to existing propositional/first-order runtime verification. We extend our tool \dejavu, experiment with new examples that take advantage of the extension.
\fi
\vspace{1ex}
\noindent {\bf conventions}
\label{sec:prelim}
In this paper, we treat and present several versions of the logic LTL. To simplify
the presentation, we prefix LTL with letters as follows:

@@

\section*{Extending Propositional LTL}
\label{sec:extending-prop-ltl}

@@

\section*{Extending First-Order LTL}

@@

\begin{equation}
\label{EFLTL}
\psi \mathit{\ where\ } r_j  ( x_j ) := 
\varphi_j (x_j) : j \in \{ 1 , \ldots , n \} 
\end{equation}
